---
title: 'ESM 262: Assignment 2'
author: "Laura Ingulsrud and Keene Morrow"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Assignment 2 - Simulating Fisheries

- Write a function that takes as input
  - a table that has prices for different fish
    - Can make up prices; they don’t have to be random
    - Can make up names for fish
  - a table that has the number caught for each fish species for each location
    - each location is in a different column
    - each fish is in a different row
- Function output will be:
  - most frequently caught fish in each location
  - total revenue for each location
  - total fisheries revenue sum
  - if user requests it graph of revenue by location and total revenue (as text)
- Place your function in an *.R file
- Generate some example data for your function; and show how this is created and used in an R markdown document 
- Submit as usual as a git link on gauchospace
- Contact Will or Naomi if you get stuck


Input to Function:

1. Table with prices for different fish
```{r}

# Create different fish (Pokemon)
fish = c("Magikarp", "Goldeen", "Seaking", "Horsea", "Seadra")

# Create prices for fish
prices = c(10, 12, 16, 20, 25)

# Create table with prices and fish
fish_prices <- data.frame(row.names = fish, prices)

```


2. Table with number caught for each fish species for each location
```{r}

# Create locations where each fish species is caught
locations = c("Vermilion City", "Cerulean City", "Pallet Town", "Pewter City", "Cinnabar Island")

# Create catch number for each location
Vermilion = c(500, 350, 290, 50, 160)
Cerulean = c(200, 430, 300, 120, 230)
Pallet = c(320, 120, 320, 35, 130)
Pewter = c(480, 190, 140, 15, 220)
Cinnabar = c(150, 400, 220, 70, 90)

# Create table with catch at each location
fish_catch <- matrix(c(Vermilion, Cerulean, Pallet, Pewter, Cinnabar), 
                     nrow = 5, ncol = 5,
                     dimnames = list(c(fish), c(locations)))


```

