---
title: 'ESM 262: Assignment 1'
author: "Laura Ingulsrud and Keene Morrow"
date: "2/24/2020"
output:
  html_document:
    df_print: paged
    theme: cosmo
---

<style>
h1.title{
    text-align: center
  }
h4.author{
    text-align: center
  }
h4.date{
    text-align: center
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Stream Power Function

**Stream power** is the rate of energy dissipation against the bed and banks of a river or stream per unit downstream length. It is given by the equation:

$\Omega =\rho gQS$

where $\Omega$ is the stream power, $\rho$ is the density of water (1000 $\frac{kg}{m^3}$), $g$ is acceleration due to gravity (9.8 $\frac{m}{s^2}$), $Q$ is discharge ($\frac{m^3}{s}$), and $S$ is the channel slope.

&nbsp;
&nbsp;

### Contract
*Input*: Discharge ($Q$) in $\frac{m^3}{s}$ and channel slope ($S$).

*Output*: Stream power ($\Omega$) in W.

*Parameters*: Density of water ($\rho$) in $\frac{g}{m^3}$ and acceleration due to gravity ($g$) in $\frac{m}{s^2}$.

### Function

```{r}

# Bring in stream power function
source("stream_power.R")

```

### Function Testing
```{r}
# Expected values
stream_power(1, 1)
stream_power(1, 1, g = 1)
stream_power(1, 1, g = 1, rho = 1)

# Error Checking
# stream_power(1)
# stream_power(1, g = 1, rho = 1)
# stream_power(1, rho = 1)
# stream_power(0, 1)
# stream_power(1, 0)
# stream_power(1, 1, g = 0, rho = 1)
# stream_power(1, 1, g = 1, rho = 0)

```

### Generate Data

```{r data1}

###### One way to generate data for the function ######

# generate discharge values from a distribution
nsample = 100
discharges = rnorm(mean=25, sd=4, n=nsample)
channelslopes = rnorm(mean=50, sd=1, n=nsample)

#create a dataframe that has rows for each model run
# columns for discharge, channel slope and stream power estimate
results = data.frame(discharge=discharges, channelslope=channelslopes, streampower=NA)

head(results)

```

```{r data2}

###### Second way to generate data for the function ######

# generate a structure to store info on our possible streams
possible_streams = data.frame(name = c("A","B","C"), discharge = c(10,20,30), channelslope = c(2,3,1))

# add to the data structure
possible_streams$power = stream_power(Q = possible_streams$discharge, S = possible_streams$channelslope)

head(possible_streams)

# show results                         
ggplot(possible_streams, aes(x=discharge, y=power, fill=channelslope))+geom_col()+labs(y="Power W", x="Discharge (m^3/s)")

```

[Graph and add some text describing how to interpret the results. Why should we care?]

