---
title: 'ESM 262: Assignment 1'
author: "Laura Ingulsrud and Keene Morrow"
date: "2/24/2020"
output:
  html_document:
    df_print: paged
    theme: cosmo
---

<style>
h1.title{
    text-align: center
  }
h4.author{
    text-align: center
  }
h4.date{
    text-align: center
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Stream Power Function

**Stream power** is the rate of energy dissipation against the bed and banks of a river or stream per unit downstream length. It is given by the equation:

$\Omega =\rho gQS$

where $\Omega$ is the stream power, $\rho$ is the density of water (1000 $\frac{kg}{m^3}$), $g$ is acceleration due to gravity (9.8 $\frac{m}{s^2}$), $Q$ is discharge ($\frac{m^3}{s}$), and $S$ is the channel slope.

&nbsp;
&nbsp;

### Contract
*Input*: Discharge ($Q$) in $\frac{m^3}{s}$ and channel slope ($S$).

*Output*: Stream power ($\Omega$) in W.

*Parameters*: Density of water ($\rho$) in $\frac{g}{m^3}$ and acceleration due to gravity ($g$) in $\frac{m}{s^2}$.

### Function

```{r}
stream_power = function(Q, S, rho=1000, g=9.8) {
  # if inputs are negative, return message "[input] must be greater than zero" using two different kinds of error checking 
  Q = ifelse((Q < 0), return("discharge must be greater than zero"), Q)
  S = ifelse((S < 0), return("channel slope must be greater than zero"), S)
  if (rho < 0) return("rho must be greater than zero")
  # calculate stream power
  result = rho * g * Q * S
  return(result)
}
```

### Function Testing
```{r}
# Expected values
str_pwr(1, 1)
str_pwr(1, 1, g = 1)
str_pwr(1, 1, g = 1, rho = 1)

# Error Checking
# str_pwr(1)
# str_pwr(1, g = 1, rho = 1)
# str_pwr(1, rho = 1)
# str_pwr(0, 1)
# str_pwr(1, 0)
# str_pwr(1, 1, g = 0, rho = 1)
# str_pwr(1, 1, g = 1, rho = 0)

```
