---
title: 'ESM 262: Assignment 1'
author: "Laura Ingulsrud and Keene Morrow"
date: "2/24/2020"
output:
  html_document:
    df_print: paged
    theme: cosmo
---

<style>
h1.title{
    text-align: center
  }
h4.author{
    text-align: center
  }
h4.date{
    text-align: center
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Stream Power Function

**Stream power** is the rate of energy dissipation against the bed and banks of a river or stream per unit downstream length. It is given by the equation:

$\Omega =\rho gQS$

where 

- $\Omega$ is the stream power
- $\rho$ is the density of water (1000 $\frac{kg}{m^3}$)
- $g$ is acceleration due to gravity (9.8 $\frac{m}{s^2}$)
- $Q$ is discharge ($\frac{m^3}{s}$)
- $S$ is the channel slope.

&nbsp;

### Function Components
*Input*: Discharge ($Q$) in $\frac{m^3}{s}$ and channel slope ($S$).

*Output*: Stream power ($\Omega$) in W.

*Parameters with Defaults*: Density of water ($\rho$) in $\frac{g}{m^3}$ (default = 1000 $\frac{g}{m^3}$) and acceleration due to gravity ($g$) in $\frac{m}{s^2}$ (default = 9.8 $\frac{m}{s^2}$).

### Function

```{r}

# Bring in stream power function
source("stream_power.R")

```

### Function Testing
```{r}
# Expected values
stream_power(1, 1)
stream_power(1, 1, g = 1)
stream_power(1, 1, g = 1, rho = 1)

# Error Checking
# stream_power(1)
# stream_power(1, g = 1, rho = 1)
# stream_power(1, rho = 1)
# stream_power(0, 1)
# stream_power(1, 0)
# stream_power(1, 1, g = 0, rho = 1)
# stream_power(1, 1, g = 1, rho = 0)

```


### Generate Data

#### Method 1
```{r data1}

###### One way to generate data for the function ######

# generate discharge values from a distribution
nsample = 1000
discharges = rnorm(mean = 25,
                   sd = 4,
                   n = nsample)

channelslopes = rnorm(mean = 50,
                      sd = 1,
                      n = nsample)

# create a dataframe that has rows for each model run
# columns for discharge, channel slope and stream power estimate
results = data.frame(discharge = discharges,
                     channelslope = channelslopes,
                     streampower = NA) %>%
  mutate(streampower = stream_power(discharge, channelslope))

head(results)

# show results                         
ggplot(results, aes(x = discharge, y = streampower, color = channelslope), alpha = 0.5) +
  geom_point() +
  labs(title = "Stream Power",
       y = "Power (W)",
       x = "Discharge (m^3/s)",
       caption = "Ingulsrud & Morrow\nESM 262 - Winter 2020") +
  theme_minimal()

```

#### Method 2
```{r data2}

###### Second way to generate data for the function ######

# generate a structure to store info on our possible streams
possible_streams = data.frame(name = c("A","B","C"),
                              discharge = seq(from = 1, to = 60.9, by = 0.1),
                              channelslope = seq(from = 30.05, to = 60, by = 0.05))

# add to the data structure
possible_streams$power = stream_power(Q = possible_streams$discharge,
                                      S = possible_streams$channelslope)

head(possible_streams)

# show results                         
ggplot(possible_streams, aes(x = discharge, y = power, color = channelslope)) +
  geom_point() +
  labs(title = "Stream Power",
       y = "Power (W)",
       x = "Discharge (m^3/s)",
       caption = "Ingulsrud & Morrow\nESM 262 - Winter 2020") +
  theme_minimal()

```

### Interpretation of Results

Power increases with discharge and channel slope.  This is consistent with our conception of energy dissipation being increased in streams of higher volumes and at steeper slopes.  A small mountain stream may have a similar power to a higher volume lowland stream due to the tradeoff between volume and slope.


